---
title: "week_10_notes"
author: "Andrew Nalundasan"
date: "11/29/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Overview video

+ Summary of the course

    + Start with obtaining TS 
        + TS must have a time trend (increasing, decreasing, cyclical/seasonal)
        + usually divide into Deterministic vs. Stochastic trends
    + Deterministic trends - need to remove the trend to get to stationarity to apply linear models
    + Stochastic trends - remove trend by taking log difference
        + Very good interpretation as it represents the growth rate
        + Make certain process is stationary by applying Dickey-Fuller Test
            + Apply ARMA or ARIMA model
            + Plot ACF and PACF and analyze spikes to determine order
    + In-Sample Evaluation
        + Both will need to check unit root MA model and not necessarily invertible
        + Check residuals
    + T-test
        + check coefs individually
    + F-test
        + check coefs jointly
    + Check residuals to determine if white noise
    + Stock price is not good for forecasting because data is usually white noise
    + Auto sample evaluations
        + Specific for forecasting
        + Focus on forecasting error
            + Split sample between estimation sample vs. prediction sample
            + Calculate forecast error by analyzing prediction sample
            + Different models will output different forecasts
            + Combined forecast will give the best forecast

+ What's next?

    + Continue independent learning with textbook
    + Extract dependency from conditional mean
        + Time dependency on conditional variance
        + Assume conditional variance
    + Expand from univariate to multi-variate TS
    + Co-integration
    

## Chapter 10 Lecture videos

+ Forecasting the long term

+ Trends

    + Deterministic
        + Trend that is linear function of time
        + Trend stationary (residucals are stationary)
            + Linear / Quadratic / Exponential / Logistic
        + Compare R2 / AIC / SIC
            + R2 <- higher the better
            + AIC / SIC <- smaller the better
        + Goal: get residuals to appear as white noise
    + Stochastic
        + aka "Random Walk" 
            1. Random Walk without Drift
                + Similar to AR(1) model
                + The shocks (spikes) never go away (get down to 0)
            2. Random Walk with Drift
              + There is a trend but unlike Deterministic, mean will come back after time infinity

+ Dickey-Fuller Test

    + Similar to t-test
    1. Random walk without Drift
        + run regression without intercept
    2. Random walk with Drift
        + run regression with intercept
    3. Random walk with Drift and Trend
        + run regression with intercept and linear time trend
        
+ Optimal Forecast

    + Can only happen when you figure out which case you're working with
    + Increase forecast horizon
    
+ Test unit root

    + Get first difference
    + Regressor is the lag(1) period of the original TS
    + t-statistic vs. critical value
        + if t-statistic < abs(critical_value) <- fail to reject Ho hypothesis

+ Residuals

    + We want residuals to be white noise (no spikes in ACF or PACF)
        + Augmented Dickey-Fuller Test
        + Get residuals to be more like white noise

+ Summary

    + Non-stationary TS, use Dickey-Fuller test to find if it has a Unit Root
        + Determine trend as Deterministic or Stochastic trend
        + Make estimations and forecasts accordingly
    
## Chapter 10 R Instructions

### Part 1 Simulation

1. Deterministic time trend

```{r}
# P258 Generate time trend
t <- c(1:50)  # generate t=1,2,...50
y <- 1 + 0.5 * t + rnorm(50)  # y=1+.5*trend + normal white noise

# plot y
ts.plot(y)

# also plot the trend
lines(.5*t)  
```

**Comments**

+ Distribution moves up and down, but always returns back to the trendline

2. Generate random walk without drift

```{r}
# P271 Generate randm walk
y <- numeric(500)  # generate a vector of 500 zeros
for (i in 2:500){  # start a for loop
  y[i] = y[i-1] + rnorm(1)  # generate random walk without drift
}  # end the loop

# plot y
ts.plot(y)

# also plot horizontal line at zero
lines(numeric(500)) 

# another way to generate random walk without drift
ts.plot(cumsum(rnorm(500))) 
```

**Comments**

+ Trend of the mean of the TS
+ See the TS random walk - does not always return back to trendline

```{r}
# P277 model for case I, showing in fig10.8a (You should run this several times to see the variation)
# generate a vector of 500 zeros
y <- numeric(500) 
for (i in 2:500){ 
  y[i] = 0.9988 * y[i-1] + rnorm(1)  # generate random walk without drift
}
# plot y
ts.plot(y) 

# also plot trend .5*t
lines(0*c(1:500))
```

**Comments**
+ coefficient close to 1
+ still has features of random walk
+ does not hug the trendline
+ it will return to trendline, but we don't know when

3. Generate random walk with drift

```{r}
# generate a vector of 500 zeros
y <- numeric(500) 
for (i in 2:500){ 
  y[i] = 0.5 + y[i-1] + rnorm(1)  # generate random walk with drift
}

# plot y
ts.plot(y) 

# also plot trend .5*t
lines(.5*c(1:500)) 
```

**Comments**

+ Drift is 0.5
+ Coefficient is 1
+ Drift and coefficient make the TS a random walk
+ Don't know when it will come back to cross the trend.

```{r}
# P277 model for case II 
# generate a vector of 500 zeros
y <- numeric(500) 
for (i in 2:500){ 
  y[i] = -0.325 + 0.9905 * y[i-1] + rnorm(1)  # generate random walk with drift
}

# plot y
ts.plot(y)

# also plot trend .5*t
lines( 0 * c(1:500))
```

**Comments**
+ TS is < 0 because we have a negative drift

4. Generate random walk with drift and time trend

```{r}
# P277 model for case III shown in fig10.8b
# generate a vector of 500 zeros
y <- numeric(500) 
for (i in 2:500){ 
  y[i] = 0.9167 + 0.0082i + 0.9830 * y[i-1] + rnorm(1)  # generate random walk with drift and a time trend
}

# plot y
ts.plot(y)

# also plot trend .5*t
lines(0 * c(1:500))  
```

### Part 2 Deterministic trend

+ How to select a model to fit the data with deterministic trend
+ Move in 2 steps

    1. Choose the order of the polynomial trend
    2. After fitting the data, remove the polynomial trend
        + Stationary / AR model to fit the residual
        
```{r}
# P266
library(readxl)
data <- read_excel("../02_raw_data/Figure10_5_MortgageDebt.xls")
debt <- ts(data$`MORTGAGE DEBT`, frequency=4, start = c(1992, 1))  # declare debt as quarterly time series
plot(debt)
```

**Comments**
+ Plot resembles deterministic trend. Does not look linear
+ This tells us we need to use a polynomial function of time to approximate the trend

```{r}
trend1 <- ts(c(1:49), frequency = 4, start = c(1992, 1))  # trend1=1,2,3...60, declared to be time-series
model <- lm(debt ~ poly(trend1, 4, raw=TRUE))  # y=b0+b1*t+b2*t^2+b3*t^31+b4*t^4+WN

# estimation output
summary(model) 
```

**Comments**
+ Coefficients of all 4 regressors are significant
+ R2 is very high
+ Need to calculate AIC and BIC

```{r}
# smaller the better
AIC(model)  # AIC
## [1] 434.5619
BIC(model)  # BIC
## [1] 445.9129
```

**Comments**

+ the smaller the better

2. Plot the original TS, calculate the fit of the residual, and plot the fitted value

```{r}
#plot debt w/o y-axis title, dashed
plot.ts(debt, ylab="", lty=2) 

# declare fit as quarterly time series
fit = ts(fitted(model), frequency = 4, start=c(1992,1)) 

# declare res as quarterly time series
res = ts(resid(model), frequency = 4, start=c(1992,1)) 

# also plot fit in red color
lines(fit, col="red", lty=2) 

# following series will be graphed on the existing plot
par(new=TRUE) 

# plot residuals w/o y-axis title, in blue color
plot.ts(res, axes=FALSE, ylab="", col="blue") 

# add the right y-axis for residuals
axis(side=4, at = pretty(range(res))) 

# legend, debt and fit will be dashed 
legend("topleft", legend=c("Actual", "Fit", "Res"), lty=c(2,2,1), col=c("black", "red", "blue")) 
```

**Comments**
+ Fitted value in red dashed line

    + This matches the original data well
    
+ Residual plot appears stationary

    + Not white noise because the TS does not fade as time moves on - it always returns back to trendline
    
+ Run ACF and PACF to choose model order

```{r}
# ACF of residuals of the 4th order polynomial trend model
acf(res)

# PACF of residuals of the 4th order polynomial trend model
pacf(res) 
```

**Comments**
+ ACF - spikes gradually move into the significance threshold
+ PACF - 2 spikes, so appears to be AR(2) to be good fit for the residuals

```{r}
# P268
# install.packages("dynlm") #install dynamic linear modelling package

# dynlm is required for using lag operators
library(dynlm) 

# AR(2) + polynomial(4) trend
model <- dynlm(debt ~ trend1 + I(trend1^2) + I(trend1^3) + I(trend1^4) + lag(debt,-1) + lag(debt,-2)) 

# estimation output
summary(model) 

AIC(model)  # AIC
## [1] 394.7132
BIC(model)  # BIC
## [1] 409.5144
```

**Comments**
+ Still maintain time trends, but add 2 more regressors
+ We see the coefficients on the 2 lags are significant (trend1) and the rest is not so significant anymore
+ Compare AIC and BIC with original model, this one is smaller, so it's better

```{r}
# P269
# plot debt w/o y-axis title, dashed
plot.ts(debt, ylab="", lty=2) 

# declare fit as quarterly time series, note that fitted values start drom 1992q3 because it is an AR(2) model
fit = ts(fitted(model), frequency = 4, start=c(1992,3)) 

# declare res as quarterly time series
res = ts(resid(model), frequency = 4, start=c(1992,1)) 

# also plot fit in red color
lines(fit, col="red", lty=2) 

# following series will be graphed on the existing plot
par(new=TRUE) 

# plot residuals w/o y-axis title, in blue color
plot.ts(res, axes=FALSE, ylab="", col="blue") 

# add the right y-axis for residuals
axis(side=4, at = pretty(range(res))) 

# legend, debt and fit will be dashed
legend("topleft", legend=c("Actual", "Fit", "Res"), lty=c(2,2,1), col=c("black", "red", "blue")) 

acf(res)
pacf(res)
```

**Comments**
+ Still difficult to judge just by the graph itself, so need to plot ACF and PACF
+ No spike in either ACF or PACF

### Part 3 Dickey-Fuller Test

+ Conduct Dickey-Fuller Test
+ Forecast using TS with Random Walk and a root

```{r}
# P281
library(readxl)

data <- read_excel("../02_raw_data/Figure10_11_Hours_USA_Spain.xls")
USA <-  ts(data$Hoursworked_USA, frequency = 1, start = 1970)
SPAIN <-  ts(data$HoursWorked_SPAIN, frequency = 1, start = 1970)

# plot the data
plot(SPAIN)
```

**Comments**
+ Does not show deterministic trend
+ This means it may have a stochastic trend
+ Critical values are different with different null hypotheses
+ Is it a random walk with or without drift, is there a time trend?

```{r}
#P283
#install.packages("urca") # Unit Root and Cointegration Tests for Time Series Data
library(urca) #use urca package, allows choosing drift and/or trend for ADF test

# augmented dickey-fuller test
x <- ur.df(SPAIN, type = "drift", lags = 0)
summary(x)

ur.df(SPAIN, type = "drift", lags = 0) 
ur.df(diff(SPAIN), type = "drift", lags = 0) 
ur.df(diff(diff(SPAIN)), type = "drift", lags = 0) 
```

**Comments**
+ Conducted linear model
+ t value of 1st lag
+ can't trust the p-value because it's not a t-distribution. it's a dickey-fuller distribution!
+ t-statistic: 0.1408 
+ f-statistic: 0.2265 
+ we fail to reject the null
+ there is a unit root

```{r}
#install.packages("dynlm") #Dynamic Linear Modelling
library(dynlm)  # use dynlm package

dSPAIN <- diff(SPAIN)
model <- dynlm(dSPAIN ~ lag(SPAIN, -1))  # regression corresponding to the above ADF test

# estimation output
summary(model) 
AIC(model)
## [1] 171.2994
BIC(model)
## [1] 175.6966

# generate residuals
res <- residuals(model) 
acf(res)
pacf(res)
```

**Comments**
+ Coefficients from dynamic model are same as Dickey Fuller test
+ Does not look like residuals are white noise because there are several spikes above the significance threshold
+ We need residuals to be white noise

```{r}
# P286
ur.df(SPAIN, type = "drift", lags = 1) 

# regression corresponding to the above ADF test
model <- dynlm(dSPAIN ~ lag(SPAIN, -1) + lag(dSPAIN, -1))  
summary(model)

AIC(model)
## [1] 134.9974
BIC(model)
## [1] 140.7334

res <- residuals(model)
acf(res)
pacf(res)
```

**Comments**
+ We fail to reject the null
+ We prefer this model to the original because AIC and BIC numbers are smaller
+ PACF - no large spikes, so the residuals are white noise
+ Now we are prepared to make forecasts

```{r}
# P287 Prepare to forecast table 10.8 
# AR(1) w/o constant for differenced series
model <-dynlm(dSPAIN ~ lag(dSPAIN,-1)+0) 
summary(model)

AIC(model)
## [1] 135.857
BIC(model)
## [1] 138.725
```

```{r}
# make 4-step forecasts
# fcast will include actual SPAIN and 4 forecasts for SPAIN
fcast <- union(SPAIN, c(1, 2, 3, 4)) 

# fcastd will include actual dSPAIN and 4 forecasts for dSPAIN
fcastd <- union(dSPAIN, c(1, 2, 3, 4)) 

for(i in 1:4){
  fcastd[32 + i] = coef(model)[1] * fcastd[31 + i]  # manually retrieving forecasts from AR(1) model for dSPAIN
  fcast[33 + i] = fcast[32 + i] + fcastd[32 + i]  # the forecast for fcast_t = fcast_t-1 +fcastd_t (note that fcastd is one period "behind", thats why the last term is fcastd[32+i], and not fcastd[33+i])
}
```


```{r}
# P288, fig 10.13
fcast <-  ts(fcast, frequency = 1, start = 1970)
ts.plot(fcast)  # plot of the forecast
lines(SPAIN, col=6)  # simultaneous plotting of the actual (original) series in magenta color
```

**Comments**
+ Magenta is the original data
+ Black color is the point forecasts

Do the same thing for the US TS

```{r}
# P289 do the same for the u.s.
plot(USA)

# ADF for USA with trend and constant
x <- ur.df(USA, type="trend", lags=1)  
summary(x)

ur.df(USA, type="trend", lags=1)  # ADF for USA with trend and constant
model <- dynlm(diff(USA) ~ lag(USA, -1) + diff(lag(USA, -1)) + trend(USA))  # corresponding regression
summary(model)
AIC(model)
## [1] 104.6664
BIC(model)
## [1] 111.8363
model <- dynlm(USA ~ lag(USA, -1) + lag(USA, -2) + trend(USA))  # AR(2) with trend and constant
summary(model)
AIC(model)
## [1] 104.6664
BIC(model)
## [1] 111.8363
```

**Comments**
+ Plot is not deterministic
+ Could be stochastic with time trend
+ This indicates it's Case 3 in DF test
+ Conduct DF test and it's Augmented DF test with lag 1
+ Results tell us what we have run
+ Dynamic Linear Model allows us to use different functions
+ Model we use here is AR(2) model

